Untitled

### 2024-12-06 analysis/paper/access-ice.qmd
Adds rmse by longitude

Unlike Arctic sea ice, which has been steadily retreating at least since the start of satellite records in the early 80s, Antarctic sea ice had a slightly increasing trend that models systematically failed to reproduce and puzzled researchers [@hobbs2016]. 
Then, in 2016 Antarctic sea ice extent dropped precipitously and has been at low and record low levels since, highlighting even more our lack of understanding of Antarctic sea ice variability and change. [ref: that CESM run with forced winds].

@@

Correctly modelling Antarctic sea ice is not only necessary for process understanding and climate projections to inform adaptation strategies. 
Accurate seasonal to subseasonal forecast are crucial for operations in and around the Antarctic continent [@desilva2020 @wagner2020]. 
However, Antarctic sea ice forecasts have been challenging both due to model biases and inherent large variability and complexity and it has lagged behind Arctic sea ice forecasts [@zampieri2019]. 
Dynamical seasonal forecasts of Summer Antarctic sea ice have been shown to perform worse than relatively simple statistical methods [@massonnet2023], which also underscores the need for better understanding of sea ice dynamics.

@@

+S2 [@wedd2022] is the Australian Bureau of Meteorology's current seasonal prediction system that replaced +S1 [@hudson2017] in October 2021. 
Both +S2 and +S1 consist on the Global Atmosphere 6.0 (GA6) [@williams2015; @waters2017], Global Land 6.0  [@best2011; @waters2017], Global Ocean 5.0 [@madec2013; @megann2014] and Global Sea Ice 6.0 [CICE; @rae2015].
The atmosphere has a N216 horizontal resolution(~60km in the mid-latitudes) with 85 levels. 
The land model uses the same horizontal grid with 4 soil levels. 
The ocean component has a 1/4º resolution with 75 vertical levels. 
The ice component --based on CICE version 4.1-- has the same resolution than the ocean and 5 sea ice categories as well as an open water category.

@@

We use NOAA/NSIDC's Climate Data Record V4 [CDR; @meier2014] as the primary verification dataset. 
Its sea ice concentration data are estimated taking the maximum value of the NASA Team [ref] and NASA Bootstrap [@comiso2023] algorithms to reduce their low concentration bias [@meier2014; @meier2021].
Both algorithms use data from the Scanning Multichannel Microwave Radiometer (SMMR) on the Nimbus-7 satellite and from the Special Sensor Microwave/Imager (SSM/I) sensors on the Defense Meteorological Satellite Program's (DMSP) -F8, -F11, and -F13 satellites. 
The data has a resolution of 25 by 25km and daily from 1978 onwards.

@@

1. The European Organisation for the Exploitation of Meteorological Satellites (EUMETSAT) Ocean and Sea Ice Satellite Application Facility (OSI SAF) [ref] is another satellite-derived sea ice concentration product based on mostly the same sensors as the NOA CDR but computed independently using different algorithms.

@@

@@

@@

```{r fig-bias}
#| fig-width: 9
#| fig-height: 6
#| fig-cap: +S2 reanalysis sea ice concentration bias compared with NSIDC sea ice concentration.

climatologies |> 
  _[model %in% c("S2", which_dataset)] |> 
  dcast(time + xgrid + ygrid ~ model, value.var = "aice") |> 
  setnames(which_dataset, "obs") |> 
  ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(aes(z = S2 - obs, fill = after_stat(level)), 
                    breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)) +
  scale_fill_divergent_discretised(paste0("Bias with respect to ", labels_models[which_dataset])) +
  geomcoord_antarctica +
  geom_antarctica_fill +
  facet_wrap(~ lubridate::month(time, label = TRUE)) +
  wide_legend +
  tag_facets()
```

@@

  geom_ribbon(aes(ymin = low, 
                  ymax = high, 
                  color = version, 
                  fill = version,
                  group = interaction(obs_dataset, version)),
              alpha = 0.1) +
  geom_line(aes(color = version, group = interaction(obs_dataset, version)), linewidth = 1) +
  scale_x_continuous(breaks = NULL) +
  labs(y = "RMSE", 
       x = "Lag") +
  scale_color_models +
  scale_fill_models +
  #   scale_color_models +
  #   scale_fill_models +
  facet_wrap(~ month, labeller = labeller(month = labels_month)) 
```

@fig-rmse shows the median and 95% range of RMSE of sea ice concentration anomalies for +S2 forecasts compared with a benchmark of persistence.
This figure uses +CDR data, but the results are nearly identical compared with ERA5 or Bootstrap.
Due to errors in the initial conditions, it is expected that a persistence forecast would be better than the model forecast at very short lags, but that the persistence forecast errors would grow faster and eventually surpass the mode forecast, at which time is statistically useful {==I've got this from CC at the ICTP summer school and makes sense, but it would be great to have a reference?=
=}.
Here the persistence errors are almost always lower than the +S2 forecast, indicating that the model doesn't have skill at any lag and in any month.
The only exception is the RMSE around February forecasted from June onwards.

@@

  scale_y_continuous(labels = labels_extent) +
  scale_x_continuous(breaks = NULL) +
  scale_color_models +
  scale_fill_models +
  labs(y = NULL, 
       x = "Lag") +
  #   scale_color_models +
  #   scale_fill_models +
  facet_wrap(~ month, labeller = labeller(month = labels_month_significant)) 
```

@@

@@

<!-- ```{r dataset_era} -->
<!-- which_dataset <- "era5" -->
<!-- ``` -->

<!-- ```{r fig-hindcast-extent-era5} -->
<!-- which_dataset <- "era5" -->
<!-- <<fig-hindcast-extent>> -->
<!-- ``` -->

<!-- ```{r fig-hindcast-extent-bt} -->
<!-- which_dataset <- "bt" -->
<!-- <<fig-hindcast-extent>> -->
<!-- ``` -->

<!-- ```{r fig-mean-growth} -->
<!-- <<fig-mean-growth>> -->
<!-- ``` -->

<!-- ```{r} -->
<!-- <<fig-bias>> -->
<!-- ``` -->

<!-- ```{r} -->
<!-- <<<fig-extent-anom>> -->
<!-- ``` -->

<!-- ```{r} -->
<!-- <<fig-rmse>> -->
<!-- ``` -->

<!-- ```{r} -->
<!-- <<fig-iiee>> -->
<!-- ``` -->

<!-- ```{r} -->
<!-- <<fig-iiee-variance>> -->
<!-- ``` -->

>>>>

Unlike Arctic sea ice, which has been steadily retreating at least since the start of satellite records in the early 80s, Antarctic sea ice had a slightly increasing trend that models systematically failed to reproduce and puzzled researchers [@hobbs2016].
Then, in 2016 Antarctic sea ice extent dropped precipitously and has been at low and record low levels since, highlighting even more our lack of understanding of Antarctic sea ice variability and change.
\[ref: that CESM run with forced winds\].

@@

Correctly modelling Antarctic sea undsen sector and the Amundsen-Bellingshausen sector (Fice is not only necessary for process understanding and climate projections to inform adaptation strategies.
Accurate seasonal to subseasonal forecast are crucial for operations in and around the Antarctic continent [@desilva2020 @wagner2020].
However, Antarctic sea ice forecasts have been challenging both due to model biases and inherent large variability and complexity and it has lagged behind Arctic sea ice forecasts [@zampieri2019].
Dynamical seasonal forecasts of Summer Antarctic sea ice have been shown to perform worse than relatively simple statistical methods [@massonnet2023], which also underscores the need for better understanding of sea ice dynamics.

@@

+S2 [@wedd2022] is the Australian Bureau of Meteorology's current seasonal prediction system that replaced +S1 [@hudson2017] in October 2021.
Both +S2 and +S1 consist on the Global Atmosphere 6.0 (GA6) [@williams2015; @waters2017], Global Land 6.0 [@best2011; @waters2017], Global Ocean 5.0 [@madec2013; @megann2014] and Global Sea Ice 6.0 [CICE; @rae2015].
The atmosphere has a N216 horizontal resolution(\~60km in the mid-latitudes) with 85 levels.
The land model uses the same horizontal grid with 4 soil levels.
The ocean component has a 1/4º resolution with 75 vertical levels.
The ice component --based on CICE version 4.1-- has the same resolution than the ocean and 5 sea ice categories as well as an open water category.

@@

We use NOAA/NSIDC's Climate Data Record V4 [CDR; @meier2014] as the primary verification dataset.
Its sea ice concentration data are estimated taking the maximum value of the NASA Team \[ref\] and NASA Bootstrap [@comiso2023] algorithms to reduce their low concentration bias [@meier2014; @meier2021].
Both algorithms use data from the Scanning Multichannel Microwave Radiometer (SMMR) on the Nimbus-7 satellite and from the Special Sensor Microwave/Imager (SSM/I) sensors on the Defense Meteorological Satellite Program's (DMSP) -F8, -F11, and -F13 satellites.
The data has a resolution of 25 by 25km and daily from 1978 onwards.

@@

2.  The European Organisation for the Exploitation of Meteorological Satellites (EUMETSAT) Ocean and Sea Ice Satellite Application Facility (OSI SAF) \[ref\] is another satellite-derived sea ice concentration product based on mostly the same sensors as the NOA CDR but computed independently using different algorithms.

@@

```{r area_daily}
area_daily <- datasets |>
  lapply(area) |>
  lapply(\(x) ReadNetCDF(x, "aice")) |>
  rbindlist(idcol = "dataset") |>
  _[, let(lon = NULL, lat = NULL)]
```

```{r area_climatology}
area_climatology <- area_daily |>  
  copy() |>
  _[aice == 0, aice := NA] |> 
  _[!(month(time) == 2 & mday(time) == 29)] |> 
  _[, time := update(time, year = 2001)] |> 
  _[, median_ci(aice), by = .(time, dataset)]

@@

```{r}
hindcast_area <- rbind(S1 = readRDS(here::here("data/derived/S1_hindcast_area.Rds")),
                       S2 = readRDS(here::here("data/derived/S2_hindcast_area.Rds")), 
                       idcol = "model") |>
  _[, model := relevel(factor(model), "S2")] |> 
  _[, lag := as.numeric(as.Date(time) - forecast_time)] |> 
  _[lag > 0] |> 
  _[!(month(forecast_time) == 1 & member == 7 & model == "S1")]  # This member is wrong
```

@@

```{r hindcast_clim}
zeropad <- function(x, n = 2) formatC(x, width = 2, flag = "0")
monnb <- function(d) { lt <- as.POSIXlt(as.Date(d, origin="1900-01-01")); 
lt$year*12 + lt$mon } 
mondf <- function(d1, d2) { monnb(d2) - monnb(d1) }

months <- 1:12
hindcast_clim <- lapply(setNames(months, months), \(month) {
  here::here(glue::glue("data/derived/hindcast_climatology/mo_aice_1981{zeropad(month)}01_mean.nc")) |> 
    ReadNetCDF("aice")
}) |> 
  rbindlist(idcol = "forecast_month") |> 
  _[, forecast_month := as.numeric(forecast_month)] |> 
  _[, forecast_time := lubridate::make_datetime(1981, forecast_month, 01)] |> 
  _[, time := lubridate::floor_date(time, "month")] |> 
  _[, lag := mondf(forecast_time, time)] |> 
  _[lag == 0] |> 
  _[, time2 := update(time, year = 2000)]

obs_clim <- CDR() |> 
  climatology() |> 
  ReadNetCDF(c(obs = "aice")) |> 
  setnames("time", "time2")

hindcast_clim <- hindcast_clim |> 
  merge(obs_clim) |> 
  _[, month := month(time)] 
```

```{r fig-bias}
#| fig-width: 9
#| fig-height: 6
#| fig-cap: 1-month lag +S2 forecast bias compared with NSIDC.

hindcast_clim |> 
  # dcast(time + xgrid + ygrid ~ model, value.var = "aice") |> 
  # setnames(which_dataset, "obs") |>
  ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(aes(z = aice - obs, fill = after_stat(level)), 
                    breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)) +
  scale_fill_divergent_discretised(paste0("Bias with respect to ",
                                          labels_models[which_dataset]),
                                   low = scales::muted("red"),
                                   high = scales::muted("blue")) +
  geomcoord_antarctica +
  geom_antarctica_fill +
  facet_wrap(~ lubridate::month(time, label = TRUE)) +
  wide_legend +
  tag_facets()
```

@@

  geom_ribbon(aes(ymin = low, 
                  ymax = high, 
                  color = version, 
                  fill = version,
                  group = interaction(obs_dataset, version)),
              alpha = 0.1) +
  geom_line(aes(color = version, group = interaction(obs_dataset, version)), linewidth = 1) +
  scale_x_continuous(breaks = NULL, expand = c(0, 0)) +
  labs(y = "RMSE", 
       x = "Lag") +
  scale_color_models +
  scale_fill_models +
  #   scale_color_models +
  #   scale_fill_models +
  facet_wrap(~ month, labeller = labeller(month = labels_month)) 
```

@fig-rmse shows the median and 95% range of RMSE of sea ice concentration anomalies for +S2 forecasts compared with a benchmark of persistence.
This figure uses +CDR data, but the results are nearly identical compared with ERA5 or Bootstrap.
Due to errors in the initial conditions, it is expected that a persistence forecast would be better than the model forecast at very short lags, but that the persistence forecast errors would grow faster and eventually surpass the mode forecast, at which time is statistically useful {==I've got this from CC at the ICTP summer school and makes sense, but it would be great to have a reference?=
=}.
Here the persistence errors are almost always lower than the +S2 forecast, indicating that the model doesn't have skill at any lag and in any month.
The only exception is the RMSE around February.

```{r}
topo_lonlat <- rcdo::cdo_topo(grid = "r180x90") |>
  rcdo::cdo_execute(options = c("-f nc")) |>
  metR::ReadNetCDF(c(z = "topo"), subset = list(lat = c(-90, -60))) |> 
  ggperiodic::periodic(lon = c(0, 360)) 
```

```{r rmse_lon}
rmse_lon <- here::here("data/derived/rmse_lon.Rds") |> 
  readRDS() |> 
  _[version != "S1"] |> 
  _[, lag := as.numeric(as.Date(time) - time_forecast)]

rmse_lon[, base := value[version == "persistence"], 
         by = .(time, lon, lag, obs_dataset)]

rmse_lon_mean <- rmse_lon |> 
  _[version != "persistence"] |>
  _[, .(dif = median(value - base, na.rm = TRUE)),
    by = .(lag, lon, version, obs_dataset, month = factor(month(time_forecast)))]
```

```{r fig-rmse_lon, fig.height=7, fig.width=10}
#| fig.cap: Meadian difference between RMSE of +S2 forecasts and persistence forecast at longitudinal bands.

rmse_lon_mean |> 
  _[obs_dataset == which_dataset] |> 
  ggperiodic::periodic(lon = c(0, 360)) |> 
  ggplot(aes(lon, lag)) +
  geom_contour_fill(aes(z = dif, fill = after_stat(level)), binwidth = 0.025)  +
  geom_contour_tanaka(aes(z = dif), range = c(0.01, 0.4),  binwidth = 0.025) +
  geom_contour2(data = topo_lonlat,
                aes(lon, scales::rescale(lat, c(1, 150)), z = z),
                breaks = 0) +
  geom_hline(data = month_from_lag(1, 1), aes(yintercept = lag),
             colour = "gray50", linewidth = 0.1) +

  geom_text(data = month_from_lag(15, 1), aes(x = 361, label = month2),
            size = 2.5, 
            hjust = 0) +

  geom_text(data = month_from_lag(1, 1), aes(x = -2, label = lag),
            size = 2.5,
            hjust = 1) +
  scale_fill_divergent_discretised() +
  scale_x_longitude(ticks = 60, expand = c(0.08, 0)) +

  scale_y_continuous(breaks = NULL, expand = c(0, 0)) +
  labs(fill = NULL, 
       y = "Lag") +
  wide_legend +
  facet_wrap(~ month, labeller = labeller(month = labels_month)) +
  theme(panel.background = element_blank()) +
  tag_facets(position = list(x = 0.1, y = 1-.02))
```

Even though +S2 is no better than persistence at forecasting pan-Antarctic sea ice concentration anomalies, the quality of the forecast might vary between regions. 
To analyse the spatial distribution of the model error, we computed ice concentration RMSE on `r uniqueN(rmse_lon_mean$lon)` meridional slices `r diff(unique(rmse_lon_mean$lon))[1]` degrees thick.
The median difference between the forecast and persistence RMSE is shown on @fig-rmse_lon, where negative values indicate that the model has lower RMSE than the benchmark.

The skill shown by +S2 at February-March forecasts is evident by a band of negative values across almost all longitudes except around and east of the Antarctic Peninsula, where most of the Summer ice is left. 
This suggests that the positive skill comes from the model correctly forecasting no ice where there is never ice. 
However, there are other regions of skilful forecasts.

Forecasts initialised on January 1^st^ ([Fig. @fig-rmse_lon] panel a) shows RMSE values below persistence at lags larger than 180 days (corresponding to July through September) in the Ross and Weddell Sea. 
These two regions are also relatively well forecasted from other months. 
Notably, May forecasts of Weddell sea ice concentration anomalies ([Fig. @fig-rmse_lon] panel e) are skillful after about 20 days and remain so until December. 
The Weddell sea region is also the region with the maximum erro --arround June irregardless of initialisation date.

@@

  scale_y_continuous(labels = labels_extent) +
  scale_x_continuous(breaks = NULL, expand = c(0, 0)) +
  scale_color_models +
  scale_fill_models +
  labs(y = NULL, 
       x = "Lag") +
  #   scale_color_models +
  #   scale_fill_models +
  facet_wrap(~ month, labeller = labeller(month = labels_month_significant)) 
```

@@

------------------------------------------------------------------------

We evaluated the sea

The ACCESS-S2 model, the Australian Bureau of Meteorology's current seasonal prediction system, shows a significant low extent bias in its sea ice extent predictions, particularly during late summer and early autumn.
This bias is attributed to a faster and longer melt season between January and March, and slower growth between March and April.
This underestimation during the minimum and early freezing season is a common issue in many seasonal-to-subseasonal (S2S) systems, suggesting potential problems with the model's thermodynamic representation.

Comparison with the previous ACCESS-S1 model reveals that while both models share the same formulation and thus similar equilibrium states, ACCESS-S1 had a smaller bias, especially at shorter forecast lead times.
This difference suggests that the data assimilation system in ACCESS-S1 was correcting for this low sea ice bias.
The analysis suggests that the data assimilation in ACCESS-S2 is only effectively influencing sea ice initial conditions from June to October, while the rest of the year, the sea ice component runs virtually free, reverting to its biased equilibrium state.

Further evaluation of ACCESS-S2's performance in forecasting sea ice anomalies reveals a poor performance even at short lead times, exhibiting much larger variability than observed.
Quantitative analysis using root mean square error (RMSE) of sea ice concentration anomalies shows that ACCESS-S2 generally lacks skill at any lead time and for any month, with its performance often worse than a simple persistence forecast.

Comparison with ACCESS-S1 using the Integrated Ice Edge Error (IIEE) metric confirms that ACCESS-S1 consistently demonstrates lower errors at short lead times, with the errors converging between the two models as the forecast progresses.
This convergence timescale, ranging from two weeks to 160 days depending on the month, is indicative of the sea ice memory of initial conditions, particularly during October to March when data assimilation has minimal influence.

Analysis of error spread further reveals that ACCESS-S1 consistently exhibits a smaller error spread at short lead times, indicating both smaller and more consistent initial condition errors.
While both models show similar ensemble spread in individual forecasts, likely due to their shared model formulation and comparable perturbation schemes, ACCESS-S2 shows a larger spread in the mean error of individual forecasts, particularly at short lead times when data assimilation is less influential on sea ice initial conditions.

------------------------------------------------------------------------

@@

<!-- ```{r dataset_era} -->

<!-- which_dataset <- "era5" -->

<!-- ``` -->

<!-- ```{r fig-hindcast-extent-era5} -->

<!-- which_dataset <- "era5" -->

<!-- <<fig-hindcast-extent>> -->

<!-- ``` -->

<!-- ```{r fig-hindcast-extent-bt} -->

<!-- which_dataset <- "bt" -->

<!-- <<fig-hindcast-extent>> -->

<!-- ``` -->

<!-- ```{r fig-mean-growth} -->

<!-- <<fig-mean-growth>> -->

<!-- ``` -->

<!-- ```{r} -->

<!-- <<fig-bias>> -->

<!-- ``` -->

<!-- ```{r} -->

<!-- <<<fig-extent-anom>> -->

<!-- ``` -->

<!-- ```{r} -->

<!-- <<fig-rmse>> -->

<!-- ``` -->

<!-- ```{r} -->

<!-- <<fig-iiee>> -->

<!-- ``` -->

<!-- ```{r} -->

<!-- <<fig-iiee-variance>> -->

<!-- ``` -->
